\section{Scientific importance}
\label{sec:science}

Sharp wave-ripples play an important role in learning and memory.\footnotemark{} The evidence for this has been accumulating over the years, and has mostly been associative / correlational. The kind of closed-loop experiments discussed in this thesis however have recently been providing more direct evidence.

\footnotetext{In this section, when we refer to ``SWR's'', we mean ``SWR's, or the mechanism that generates them''. -- The SWR pattern in the LFP may not be more than an epiphenomenon. Additionally, when we refer to ``memory'', we more specifically mean so called \emph{declarative} memory. This excludes conditioned and ``instinctive'' emotional responses, ``muscle memory'' (habits and motor skills), and habituation or sensitization of the senses.}




\subsection{Overview of evidence}

We first summarize what is known about the importance of sharp wave-ripples.
% Most of these claims are supported by both clinical studies and animal experiments.
In the following subsections, we expand on each of these points.

\begin{enumerate}
\item The hippocampus -- where SWR's are recorded from -- is necessary for consolidating memories.
\item Memory consolidation happens mostly during sleep. SWR's also occur mostly during sleep.
\item The ideal firing frequency for strengthening synapses is the same as the ripple frequency.
\item Awake neural firing patterns are later replayed in the hippocampus. This replay occurs mostly during SWR's.
\item Forcibly silencing the hippocampus during ripples after a new task is learned deteriorates subsequent performance on the task. Silencing the hippocampus outside ripples has no such effect.
\end{enumerate}

The last two points have only been convincingly demonstrated in rats, but are likely true for other mammals as well.




\subsection{Memory consolidation}

SWR's are thought to be pertinent mainly in the process of \emph{memory consolidation}, where certain short-term memories are stabilized into long-term memories. Single-cell recordings in primates have hinted that short-term memories exist as positive feedback loops of firing neurons in the neocortex (``reverberations'', or persistent activity). Long-term memories on the other hand are likely to exist as strenghtened or newly built physical connections between neurons. \cite{Kandel2013,Bear2015}

The case of Henry Molaison (known as patient H.M. up until his death in 2008) illustrates the importance of memory consolidation, and of the role of the hippocampus in this process. In an attempt to cure his epilepsy, most of Molaison's hippocampi and the adjacent entorhinal cortices\footnote{Main input to the hippocampus, relaying deeply processed sensory information from the rest of the neocortex.} were surgically removed (leaving the rest of the neocortex intact). Before the surgery, Molaison had no memory problems. After the surgery however -- although his epileptic seizures decreased -- he could not form new long-term memories.\footnotemark{} Curiously though, Molaison still had both intact short-term memory (remembering new information for up to minutes after), and largely intact long-term memory for events that happened before the surgery. These symptoms are consistent with a model of memory consolidation where both short-term and long-term memories are stored and retrieved by the neocortex, but where the hippocampus is needed to convert short-term into new long-term memories. \cite{Kandel2013}

\footnotetext{For example, the doctors working with Molaison had to re-introduce themselves on every occasion they saw him \cite{Bear2015}.}


% systems vs synaptic consolidation
% sleep, two stage-model


\subsection{Long-term potentiation}





\subsection{Replay}

Quite recently, a discovery was made in the hippocampus that yields evidence for the importance of sharp wave-ripples at a more detailed and mechanistic level. This discovery, called \emph{replay}, was made through simultaneous recordings of the spiking activity of multiple individual neurons in the hippocampus.\footnotemark{}

\footnotetext{Such recordings are classically made with tetrodes -- four tightly wound electrodes. More recently, multi-electrode neural probes are also being used. Each electrode picks up the spikes from multiple nearby neurons; and each neuron's spikes are visible on multiple tetrodes. Each recorded spike is assigned to an individual neuron in a process called \emph{spike sorting}.}

When a rat is placed in a maze, there is a good chance that a given hippocampal neuron is a so called \emph{place cell}, meaning that it only fires around a certain location in the maze.\footnotemark{}. When the rat walks around the maze, a set of place cells will therefore fire in sequence, reflecting the path that the rat traversed.

\footnotetext{The area where it fires is called the neuron's \emph{place field}. To be precise, the neuron can fire anywhere in the maze, but has an increased firing \emph{rate} in its place field. Finding the rat's position based on neuron firing rates is called \emph{neural population decoding}, and is typically performed using a Bayesian framework to invert the empirical place field encodings \cite{Kloosterman2012}.}

When the rat later rests, these place cell firing sequences occur again, even though the rat stays in the same place. These sequence reactivations are called \emph{replay}. They are compressed in time versus the original sequences, and, importantly, occur mostly during sharp wave-ripples.

Replay is thus a quite literal ...





\subsection{Ripple disruption}

Even more recently, ...

% \cite{Ego-Stengel2009,Girardeau2009,Jadhav2012,Girardeau2014,Talakoub2016}
