\chapter*{Preface}
\label{sec:preface}
\addcontentsline{toc}{chapter}{\nameref{sec:preface}}

% I am priviliged.

% insert 'especially' if more text before this
I am very grateful that I got the opportunity to do my master's thesis at NERF\footnote{Neuro-Electronics Research Flanders. A collaborative research initiative between the university of Leuven, the Flemish life sciences institute VIB, and imec (a ``boutique, not-for-profit microelectronics shop'', in the \href{https://www.economist.com/science-and-technology/2017/11/09/a-new-nerve-cell-monitor-will-help-those-studying-brains}{words} of The Economist).}, where I was introduced to the wonderful world of circuit-level neuroscience.\footnote{I watched the fluorescent neural fireworks through a laser confocal microscope; I heard live the crackling firing of neurons in a sleeping brain; and I saw beautiful brain tissue photographs where virally infected cells fluoresced in bright colours. I learned about animal surgery, 3D printing, behavioural experiments, cluster and GPU computing, and the manufacture of bespoke, electromechanical scientific tools.}
I want to thank Alexander Bertrand and Fabian Kloosterman, whose collaboration made this possible. They, and my thesis supervisors Jasper Wouters and Davide Ciliberti, spent many hours on very helpful feedback, guidance, and support.

I also want to thank Jo\~ao Couto, \c{C}a\u{g}atay Ayd{\i}n, and Luis Hoffman for setting up the NERF computing cluster, writing a comprehensive usage guide for it, and helping me to get started with it. I thank Luis Hoffman specifically for his magnanimous help with various imec networking issues.

Finally, a big thanks to my family and friends for the continued support throughout the thesis work. You know who y'all are.

 % todo: make link a proper reference
