\chapter*{Preface}
\label{sec:preface}
\addcontentsline{toc}{chapter}{\nameref{sec:preface}}

% I am priviliged.

% insert 'especially' if more text before this
I am very grateful that I got the opportunity to do my master's thesis at NERF\footnotemark, where I was introduced to the wonderful world of circuit-level neuroscience. I watched the fluorescent fireworks of a thousand individual  neurons through a laser confocal microscope; I heard live the crackling firing of neurons deep in the brain of a sleeping rat; and I saw beautiful brain tissue photographs where virally infected cells fluoresced in many bright colours. I learned about animal surgery, 3D printing, behavioural experiments, cluster and GPU computing, and the manufacture of bespoke, electromechanical scientific tools.

For all this I want to thank Alexander Bertrand and Fabian Kloosterman. Their collaboration made this possible.

I also want to thank Jo\~ao Couto, \c{C}a\u{g}atay Ayd{\i}n, and Luis Hoffman for setting up the NERF computing cluster, writing a comprehensive usage guide for it, and helping me to get started with it. I thank Luis Hoffman specifically for his magnanimous help with various imec networking issues.


\footnotetext{Neuro-Electronics Research Flanders. A collaborative research initiative between the university of Leuven, the Flemish life sciences institute VIB, and imec (a ``boutique, not-for-profit microelectronics shop'', in the \href{https://www.economist.com/science-and-technology/2017/11/09/a-new-nerve-cell-monitor-will-help-those-studying-brains}{words} of The Economist).} % todo: make link a proper reference
